\section{Amazon: product recommendation graph}

The dataset consists of a product recommendation graph for an online catalog (Amazon) (3). It comprises a list of products along with associated metadata, review details, and relationships such as \textit{also purchased} and \textit{also viewed}. This dataset
\footnote{The processed dataset can be downloaded from \url{https://drive.google.com/file/d/13A9DCImmOZPFI6TyHZRUpaZcMQb3Rq0V/view?usp=sharing}.}was extracted from the SKB (Semi-structured Knowledge Base) of the Stark project \cite{stark2024}, which provides three SKBs, one of which is the Amazon catalogue available at \url{https://stark.stanford.edu/dataset_amazon.html}.

The graph is archived in a ZIP file with a total size of 208 MB and contains slightly fewer than 1 million nodes and 4 million edges. In its uncompressed state, the nodes and edges files require 663 MB and 94.9 MB, respectively.

Nodes represent either products or categories. The edges connect products to their respective categories and further link products together based on the following user interactions:
\begin{itemize}
    \item Users who have viewed both product pages.
    \item Users who have purchased both products.
\end{itemize}

\paragraph{Node Properties (all defined for products, with only a subset applicable to categories):}
\begin{itemize}
    \item \emph{ID}: Unique identifier.
    \item \emph{Type}: Specifies whether the node represents a product or a category.
    \item \emph{Asin}: An alternative identifier, valid only for products.
    \item \emph{Title}: The name of the product in the Amazon catalogue, or the name of the category if the node is of type category.
    \item \emph{Review\_score}: The average review rating available on the website (a float ranging from 1 to 5).
    \item \emph{Price}: The product price in dollars, or 0 if unspecified.
    \item \emph{Brand}: The product's brand.
    \item \emph{Feature}: A list of tags associated with the product on the Amazon catalogue.
    \item \emph{Rank}: A string that specifies the ranking of the product within its corresponding category.
    \item \emph{Description}: A text description providing additional details about the product.
\end{itemize}

\paragraph{Edge Properties:}
\begin{itemize}
    \item \emph{Source}: The identifier of the source node.
    \item \emph{Type}: The type of relationship (can be ALSO\_VIEWED, ALSO\_BOUGHT, HAS\_CATEGORY).
    \item \emph{Target}: The identifier of the target node.
\end{itemize}